---
title: "bln_ess_groundwater"
output: rmarkdown::html_vignette
bibliography: references.bib
csl: nmi-csl-nmi-v1_11.csl
vignette: >
  %\VignetteIndexEntry{bln_ess_groundwater}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}

# required packages
require(data.table); require(BLN); require(ggplot2);require(patchwork)

# set some default options for the html page
knitr::opts_chunk$set(echo = FALSE,
                      collapse = TRUE,
                      comment = "#>")
options(rmarkdown.html_vignette.check_title = FALSE)
setDTthreads(1)

```

```{r setup}
library(BLN)

ggplot2::theme_set(
  theme_bw() +
  theme(
    plot.title = element_text(size=10),
    legend.text = element_text(size=10),
    axis.text = element_text(size=10, colour = 'black'),
    legend.position="inside",
    legend.position.inside = c(0.7,0.2),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank()
  )
)
```

## Introduction
Groundwater is used as source of drinking water, irrigation, and supply habitats
with water. Groundwater must continuously be recharged by precipitation which
infiltrates the soil. Soil plays a pivotal role in ensuring that there is ample
clean groundwater. Soils may buffer groundwater recharge, filter out nutrients,
and breakdown (organic) contaminants such as pesticides.

It is important to monitor the soil functions that support groundwater ecosystem
services. Because agricultural soils cover large areas, and may be subject to 
operations which hinder soil functions in support of groundwater such as the use
of large machines that may compact the soil, and the use of manure, fertilisers,
and pesticides.

The BLN evaluates the following soil functions:

* groundwater recharge
* buffering of nitrogen in soil water
* buffering of pesticides

## Groundwater recharge
Groundwater recharge is evaluated with the `bln_wat_groundwater_recharge` and
`bln_bbwp_bw` functions. These evaluations are also used separately in the OBIC
and BBWP packages [@verweij_obic_2023; @bbwp2023].

The first function, `bln_wat_groundwater_recharge`, evaluates how well water can
infiltrate and seep down to the groundwater. This is affected by the soil's 
cultivation (B_LU_BRP), the risk of soil compaction (B_SC_WENR), groundwater level
(B_GWL_CLASS), presence of drains, soil texture, and the use of green manure.

The second function, `bln_bbwp_bw`, evaluates how well a soil can hold onto water
such that it has sufficient time to infiltrate. This also depends on the 
cultivation, groundwater level and soil texture but is unaffected by soil
compaction risk. Instead, this function uses considers whether the soil is 
located in a drought-prone area and the HELP code corresponding to the fields 
location [@van_Bakel2005]. These HELP codes relate to the water needs of types
of landuse.

See examples of these BLN functions below.

```{r groundwater recharge function demo, echo=TRUE}
water_recharge <- bln_wat_groundwater_recharge(
      ID = 15,
      B_LU_BRP = 259,
      B_SC_WENR = 3,
      B_GWL_CLASS = 'GtVI',
      B_DRAIN = TRUE,
      A_CLAY_MI = 20,
      A_SAND_MI = 65,
      A_SILT_MI = 10,
      A_SOM_LOI = 5,
      M_GREEN = FALSE
      )

water_buffering <- bln_bbwp_bw(
      ID = 15,
      B_LU_BRP = 259,
      B_HELP_WENR = 'Mn25C',
      B_GWL_CLASS = 'GtVI',
      B_AREA_DROUGHT = TRUE,
      A_CLAY_MI = 20,
      A_SAND_MI = 65,
      A_SILT_MI = 10,
      A_SOM_LOI = 5,
      penalty = TRUE
      )
```

Below we illustrate how these indicators react to changes in soil texture,
groundwater level, and crop type.

```{r groundwater recharge sensitivty figures, fig.width=8, fig.height=10, cache=TRUE}
# create table with variety in texture ====
dt.tex <- data.table(
  ID = 15,
  B_LU_BRP = rep(259,20),
  B_SC_WENR = rep(3,20),
  B_GWL_CLASS = rep('GtIII', 20),
  B_DRAIN = rep(TRUE,20),
  A_CLAY_MI = rep(seq(1,46,5), 2),
  A_SAND_MI = rep(seq(84,39,-5),2),
  A_SILT_MI = rep(10,20),
  A_SOM_LOI = c(rep(0.5,10), rep(5,10)),
  M_GREEN = rep(FALSE,20),
  B_HELP_WENR = rep('Mn25C',20),
  B_AREA_DROUGHT = rep(TRUE,20)
)

dt.tex[,i_gw_gwr := bln_wat_groundwater_recharge(ID,B_LU_BRP, B_SC_WENR, B_GWL_CLASS,
                                                 B_DRAIN, A_CLAY_MI,A_SAND_MI,
                                                 A_SILT_MI, A_SOM_LOI, M_GREEN)]

dt.tex[,i_gw_wb := bln_bbwp_bw(ID,B_LU_BRP, B_HELP_WENR, B_GWL_CLASS, B_AREA_DROUGHT,
                               A_CLAY_MI, A_SAND_MI, A_SILT_MI, A_SOM_LOI)]

# make A_SOM_LOI factor for easier plotting
dt.tex[,A_SOM_LOI := as.factor(A_SOM_LOI)]

p.recharge <- ggplot(
  data = dt.tex,
  mapping = aes(x = A_CLAY_MI, y = i_gw_gwr, col = A_SOM_LOI)
  ) +
  geom_point(size = 3) +
  geom_line() +
  ylim(0,1) +
  scale_color_viridis_d() +
  ylab("Groundwater recharge score") +
  xlab("Percentage clay") +
  guides(colour = guide_legend("% organic matter")) +
  ggtitle("Effect of soil texture on groundwater recharge.")

p.buffer <- ggplot(
  data = dt.tex,
  mapping = aes(x = A_CLAY_MI, y = i_gw_wb, col = A_SOM_LOI)
) +
  geom_point(size = 3) +
  geom_line() +
  ylim(0,1) +
  scale_color_viridis_d() +
  ylab("Groundwater buffering score") +
  xlab("Percentage clay") +
  guides(colour = guide_legend("% organic matter")) +
  ggtitle("Effect of soil texture on groundwater buffering.")

# create table with variety in groundwaterclass ====
dt.gwl <-  data.table(
  ID = 15,
  B_LU_BRP = rep(259, 16),
  B_SC_WENR = rep(3, 16),
  B_GWL_CLASS = rep(c('GtI', 'GtII', 'GtIII', 'GtIV', 'GtV', 'GtVI', 'GtVII', 'GtVIII'), 2),
  B_DRAIN = c(rep(FALSE, 8), rep(TRUE, 8)),
  A_CLAY_MI = rep(15, 16),
  A_SAND_MI = rep(73, 16),
  A_SILT_MI = rep(10, 16),
  A_SOM_LOI = rep(2, 16),
  M_GREEN = rep(FALSE, 16),
  B_HELP_WENR = rep('Mn25C', 16),
  B_AREA_DROUGHT = rep(TRUE, 16)
)

dt.gwl[,i_gw_gwr := bln_wat_groundwater_recharge(ID,B_LU_BRP, B_SC_WENR, B_GWL_CLASS,
                                                 B_DRAIN, A_CLAY_MI,A_SAND_MI,
                                                 A_SILT_MI, A_SOM_LOI, M_GREEN)]

dt.gwl[,i_gw_wb := bln_bbwp_bw(ID,B_LU_BRP, B_HELP_WENR, B_GWL_CLASS, B_AREA_DROUGHT,
                               A_CLAY_MI, A_SAND_MI, A_SILT_MI, A_SOM_LOI)]

p.gwl <-ggplot(
  data = dt.gwl,
  mapping = aes(x = B_GWL_CLASS, y = i_gw_gwr, fill = B_DRAIN, group = B_DRAIN)
) +
  geom_col(position = 'dodge') +
  ylim(0, 1) +
  scale_fill_viridis_d() +
  ylab('Groundwater recharge score') +
  xlab('Groundwater class') +
  guides(fill = guide_legend(title = 'Drain presence')) +
  ggtitle("Effect of groundwaterclass and drainage on\ngroundwater recharge score")

p.gwlbuf <- ggplot(
  data = dt.gwl,
  mapping = aes(x = B_GWL_CLASS, y = i_gw_wb, fill = B_DRAIN, group = B_DRAIN)
) +
  geom_col(position = 'dodge') +
  ylim(0, 1) +
  scale_fill_viridis_d() +
  ylab('Groundwater recharge score') +
  xlab('Groundwater class') +
  guides(fill = guide_legend(title = 'Drain presence')) +
  ggtitle("Effect of groundwaterclass and drainage on\nsoilwater buffering score")

# make crop table ====
dt.crop <- data.table(
    ID = 15,
  B_LU_BRP = c(256, 259, 2014, 265),
  B_SC_WENR = rep(3, 4),
  B_GWL_CLASS = rep('GtIV', 4),
  B_DRAIN = rep(FALSE, 4),
  A_CLAY_MI = rep(15, 4),
  A_SAND_MI = rep(73, 4),
  A_SILT_MI = rep(10, 4),
  A_SOM_LOI = rep(2, 4),
  M_GREEN = rep(FALSE, 4),
  B_HELP_WENR = rep('Mn25C', 4),
  B_AREA_DROUGHT = rep(TRUE, 4),
  Crop = factor(c("sugarbeet", "silage maize", 'ware potato', 'permanent grassland'),
                levels = c("sugarbeet", "silage maize", "permanent grassland", "ware potato"))
)

dt.crop[,i_gw_gwr := bln_wat_groundwater_recharge(ID,B_LU_BRP, B_SC_WENR, B_GWL_CLASS,
                                                 B_DRAIN, A_CLAY_MI,A_SAND_MI,
                                                 A_SILT_MI, A_SOM_LOI, M_GREEN)]

dt.crop[,i_gw_wb := bln_bbwp_bw(ID,B_LU_BRP, B_HELP_WENR, B_GWL_CLASS, B_AREA_DROUGHT,
                               A_CLAY_MI, A_SAND_MI, A_SILT_MI, A_SOM_LOI)]

p.cropr <- ggplot(
  data = dt.crop,
  mapping = aes(x = Crop, y = i_gw_gwr, group = Crop, fill = Crop)
) +
  geom_col(position = 'dodge') +
  ylim(0, 1) +
  ylab('Groundwater recharge score') +
  ggtitle("Effect of crop on recharge score") +
  theme(legend.position = "none") +
  scale_fill_viridis_d() +
  coord_flip()

p.cropb <- ggplot(
  data = dt.crop,
  mapping = aes(x = Crop, y = i_gw_wb, group = Crop, fill = Crop)
) +
  geom_col(position = 'dodge') +
  ylim(0, 1) +
  ylab('Groundwater buffering score') +
  ggtitle("Effect of crop on soilwaterbuffering score") +
  theme(legend.position = "none") +
  scale_fill_viridis_d() +
  coord_flip()
  
(p.recharge + p.buffer) /
  (p.gwl + p.gwlbuf)/
  (p.cropr / p.cropb)
```

## Nutrient buffering
Nutrient buffering is the capacity of the soil to retain nutrients such that
they remain available to plants and avoid leaching to groundwater.
In the BLN, this is evaluated with the functions `bln_wat_nretention_gw`,
`bln_bbwp_ngw`, and `bln_wat_nrisk_gw`. The first two assess nitrogen retention
and the third scores the risk of nitrogen leaching.These evaluations are also 
used separately in the OBIC and BBWP packages [@verweij_obic_2023; @bbwp2023].

Retention of nitrogen by the soil is mainly governed by soil type
and soil nutrient contents. Soils more vulnerable to leaching such as sandy ones,
have higher risks and score lower if all else stays the same. See the differences
for sand, peat, and clay when soil organic matter = 1.7%, total nitrogen = 2500 
mg N kg^-1^, and the C:N ratio is 10.

```{r nitrogen retention demo soiltype, fig.width=7, fig.height=3, cache=TRUE}
# make data
dt <- data.table(
  ID = 15,
  B_LU_BRP = rep(259,3),
  B_SOILTYPE_AGR = c('dekzand', 'veen', 'zeeklei'),
  B_GWL_CLASS = rep('GtIV',3),
  B_AER_CBS = rep('LG06',3),
  B_SC_WENR = rep(11,3),
  B_FERT_NORM_FR = rep(1,3),
  B_N_RT = NA_real_,
  B_N_RT_SD = NA_real_,
  B_GWP = FALSE,
  A_SOM_LOI = rep(1.7, 3),
  A_N_RT = rep(2500, 3),
  A_CN_FR = rep(10, 3),
  penalty = TRUE,
  A_CLAY_MI = rep(10,3),
  A_SAND_MI = rep(70, 3),
  A_SILT_MI = rep(20, 3),
  A_P_AL = rep(25, 3),
  A_P_WA = rep(24, 3),
  A_P_CC = rep(1.5, 3),
  A_PH_CC = rep(5.6, 3),
  A_CEC_CO = rep(122, 3),
  A_K_CC = rep(95, 3),
  A_K_CO_PO = rep(1.6, 3),
  M_GREEN = FALSE
)

# calc N retention
dt[, i_gw_nret := bln_wat_nretention_gw(ID, B_LU_BRP, B_SOILTYPE_AGR, B_AER_CBS,
                                        B_GWL_CLASS, A_SOM_LOI, A_N_RT,A_CN_FR)]
dt[, i_gw_ngw := bln_bbwp_ngw(ID,B_LU_BRP, B_SOILTYPE_AGR, B_SC_WENR, B_AER_CBS,
                              B_GWP,B_GWL_CLASS, A_SOM_LOI, A_N_RT, B_N_RT,
                              B_N_RT_SD,penalty = TRUE)]
dt[, i_gw_nlea := bln_wat_nrisk_gw(ID,B_LU_BRP,B_SOILTYPE_AGR,B_AER_CBS,
                                   B_GWL_CLASS,B_SC_WENR,B_FERT_NORM_FR = B_FERT_NORM_FR,
                                   A_CLAY_MI,A_SAND_MI, A_SILT_MI, A_SOM_LOI,A_P_AL, A_P_WA, A_P_CC,
                                   A_PH_CC, A_CEC_CO,A_K_CO_PO, A_K_CC,
                                   M_GREEN = FALSE)]

# translate
dt[B_SOILTYPE_AGR == 'dekzand', soiltype := "sand"]
dt[B_SOILTYPE_AGR == 'veen', soiltype := "peat"]
dt[B_SOILTYPE_AGR == 'zeeklei', soiltype := "clay"]

p.nret <- ggplot(data = dt,
                 mapping = aes(x = factor(soiltype), y = i_gw_nret)) +
  geom_col() +
  ylim(0, 1) +
  xlab('Soiltype') +
  ylab("Score") +
  ggtitle("N retention")

p.nbuf <- ggplot(
  data = dt,
  mapping = aes(x = factor(soiltype), y = i_gw_nret)
) +
  geom_col() +
  ylim(0, 1) +
  xlab('Soiltype') +
  ylab("Score") +
  ggtitle("N buffering")

p.nleach <- ggplot(
  data = dt,
  mapping = aes(x = factor(soiltype), y = i_gw_nlea)
) +
  geom_col() +
  ylim(0, 1) +
  xlab('Soiltype') +
  ylab("Score") +
  ggtitle("N leaching")

p.nret + p.nbuf + p.nleach

```

The soil organic matter content, C:N ratio and total soil nitrogen can also affect
the N retention score (i_gw_nret) and risk of N leaching to groundwater 
(i_gw_ngw and i_gw_nlea).

```{r N retention demo N content, fig.height=4, fig.width=7, chache = TRUE}
# Organic matter demo =====
# make data
dt <- data.table(
  ID = 15,
  B_LU_BRP = rep(259, 8),
  B_SOILTYPE_AGR = rep('dekzand',  8),
  B_GWL_CLASS = rep('GtIV',  8),
  B_AER_CBS = rep('LG06',  8),
  B_SC_WENR = rep(11,  8),
  B_FERT_NORM_FR = rep(1,  8),
  B_N_RT = NA_real_,
  B_N_RT_SD = NA_real_,
  B_GWP = FALSE,
  A_SOM_LOI = seq(0.5, 4, 0.5),
  A_N_RT = rep(2500,  8),
  A_CN_FR = rep(10,  8),
  penalty = TRUE,
  A_CLAY_MI = rep(10, 8),
  A_SAND_MI = rep(70, 8),
  A_SILT_MI = rep(20, 8),
  A_P_AL = rep(25, 8),
  A_P_WA = rep(24, 8),
  A_P_CC = rep(1.5, 8),
  A_PH_CC = rep(5.6, 8),
  A_CEC_CO = rep(122, 8),
  A_K_CC = rep(95, 8),
  A_K_CO_PO = rep(1.6, 8),
  M_GREEN = FALSE
)

# calc N retention
dt[, i_gw_nret := bln_wat_nretention_gw(ID, B_LU_BRP, B_SOILTYPE_AGR, B_AER_CBS,
                                        B_GWL_CLASS, A_SOM_LOI, A_N_RT,A_CN_FR)]
dt[, i_gw_ngw := bln_bbwp_ngw(ID,B_LU_BRP, B_SOILTYPE_AGR, B_SC_WENR, B_AER_CBS,
                              B_GWP,B_GWL_CLASS, A_SOM_LOI, A_N_RT, B_N_RT,
                              B_N_RT_SD,penalty = TRUE)]
dt[, i_gw_nlea := bln_wat_nrisk_gw(ID,B_LU_BRP,B_SOILTYPE_AGR,B_AER_CBS,
                                   B_GWL_CLASS,B_SC_WENR,B_FERT_NORM_FR = B_FERT_NORM_FR,
                                   A_CLAY_MI,A_SAND_MI, A_SILT_MI, A_SOM_LOI,A_P_AL, A_P_WA, A_P_CC,
                                   A_PH_CC, A_CEC_CO,A_K_CO_PO, A_K_CC,
                                   M_GREEN = FALSE)]

dtm <- melt(dt, value.name = 'score',
            variable.name = 'indicator',
            measure.vars = c("i_gw_nret", "i_gw_ngw", "i_gw_nlea"))
dtm[, indicator := factor(indicator, levels = c("i_gw_nret", "i_gw_nlea", "i_gw_ngw"))] # use factors with levels so each indicator is the same colour in each plot

p.som <- ggplot(
  data = dtm,
  mapping = aes(x = A_SOM_LOI, y = score, colour = indicator)
) +
  geom_point() +
  geom_line() +
  ylim(0, 1) +
  xlab("Percentage soil organic matter") +
  ylab("N retention score") +
  scale_colour_viridis_d() +
  theme(legend.position = "right")

# variying A_N_RT ====
dt <- data.table(
  ID = 15,
  B_LU_BRP = rep(259, 8),
  B_SOILTYPE_AGR = rep('dekzand',  8),
  B_GWL_CLASS = rep('GtIV',  8),
  B_AER_CBS = rep('LG06',  8),
  B_SC_WENR = rep(11,  8),
  B_FERT_NORM_FR = rep(1,  8),
  B_N_RT = NA_real_,
  B_N_RT_SD = NA_real_,
  B_GWP = FALSE,
  A_SOM_LOI = rep(1.5, 8),
  A_N_RT = seq(500, 16000,  2000),
  A_CN_FR = rep(10,  8),
  penalty = TRUE
)

# calc N retention
dt[, i_gw_nret := bln_wat_nretention_gw(ID, B_LU_BRP, B_SOILTYPE_AGR, B_AER_CBS,
                                        B_GWL_CLASS, A_SOM_LOI, A_N_RT,A_CN_FR)]
dt[, i_gw_ngw := bln_bbwp_ngw(ID,B_LU_BRP, B_SOILTYPE_AGR, B_SC_WENR, B_AER_CBS,
                              B_GWP,B_GWL_CLASS, A_SOM_LOI, A_N_RT, B_N_RT,
                              B_N_RT_SD,penalty = TRUE)]

dtm <- melt(dt, value.name = 'score',
            variable.name = 'indicator',
            measure.vars = c("i_gw_nret", "i_gw_ngw"))
dtm[, indicator := factor(indicator, levels = c("i_gw_nret", "i_gw_ngw"))]

p.nrt <- ggplot(
  data = dtm,
  mapping = aes(x = A_N_RT, y = score, col = indicator)
) +
  geom_point() +
  geom_line() +
  ylim(0, 1) +
  xlab(expression("Total nitrogen mg kg"^-1~ "soil")) +
  ylab("Score") +
  scale_colour_viridis_d() +
  theme(legend.position = "right")

# variying CN ratio ====
dt <- data.table(
  ID = 15,
  B_LU_BRP = rep(259, 16),
  B_SOILTYPE_AGR = rep('dekzand',  16),
  B_GWL_CLASS = rep('GtIV',  16),
  B_AER_CBS = rep('LG06',  16),
  B_SC_WENR = rep(11,  16),
  B_FERT_NORM_FR = rep(1,  16),
  B_N_RT = NA_real_,
  B_N_RT_SD = NA_real_,
  B_GWP = FALSE,
  A_SOM_LOI = rep(1.5, 16),
  A_N_RT = rep(7000, 16),
  A_CN_FR = seq(5, 12.5, 0.5),
  penalty = TRUE
)

# calc N retention
dt[, i_gw_nret := bln_wat_nretention_gw(ID, B_LU_BRP, B_SOILTYPE_AGR, B_AER_CBS,
                                        B_GWL_CLASS, A_SOM_LOI, A_N_RT,A_CN_FR)]
p.cn <- ggplot(
  data = dt,
  mapping = aes(x = A_CN_FR, y = i_gw_nret, colour = "#440154")
) +
  geom_point() +
  geom_line() +
  ylim(0, 1) +
  xlab("C:N ratio") +
  ylab("N retention score") +
  scale_colour_viridis_d() +
  theme(legend.position = 'none')

(p.cn+ p.som )/ p.nrt 
```

The risk of leaching nitrogen to groundwater (i_gw_nlea) is primarily determined
by the soil type (as seen above) and crop (see below). The function used to 
calculate this indicator corrects for situations where supply of P and K and the 
pH are optimal. In such situations, the risk of leaching is reduced as crops are
likely to take up more nitrogen under ideal conditions.

```{r, demo N leaching for different crops, fig.height=4, fig.width=7}
# make data
dt <- data.table(
  ID = 15,
  B_LU_BRP = rep(c(256, 259, 265, 2014), 3),
  B_SOILTYPE_AGR = c(rep('dekzand', 4), rep('zeeklei', 4), rep("veen", 4)),
  B_GWL_CLASS = rep('GtIV',  12),
  B_AER_CBS = rep('LG06',  12),
  B_SC_WENR = rep(11,  12),
  B_FERT_NORM_FR = rep(1,  12),
  B_N_RT = NA_real_,
  B_N_RT_SD = NA_real_,
  B_GWP = FALSE,
  A_SOM_LOI = rep(0.5, 12),
  A_N_RT = rep(2500,  12),
  A_CN_FR = rep(10,  12),
  penalty = TRUE,
  A_CLAY_MI = rep(10, 12),
  A_SAND_MI = rep(70, 12),
  A_SILT_MI = rep(20, 12),
  A_P_AL = rep(25, 12),
  A_P_WA = rep(24, 12),
  A_P_CC = rep(1.5, 12),
  A_PH_CC = rep(5.6, 12),
  A_CEC_CO = rep(122, 12),
  A_K_CC = rep(95, 12),
  A_K_CO_PO = rep(1.6, 12),
  M_GREEN = FALSE
)

# calc N retention
dt[, i_gw_nret := bln_wat_nretention_gw(ID, B_LU_BRP, B_SOILTYPE_AGR, B_AER_CBS,
                                        B_GWL_CLASS, A_SOM_LOI, A_N_RT,A_CN_FR)]
dt[, i_gw_ngw := bln_bbwp_ngw(ID,B_LU_BRP, B_SOILTYPE_AGR, B_SC_WENR, B_AER_CBS,
                              B_GWP,B_GWL_CLASS, A_SOM_LOI, A_N_RT, B_N_RT,
                              B_N_RT_SD,penalty = TRUE)]
dt[, i_gw_nlea := bln_wat_nrisk_gw(ID,B_LU_BRP,B_SOILTYPE_AGR,B_AER_CBS,
                                   B_GWL_CLASS,B_SC_WENR,B_FERT_NORM_FR = B_FERT_NORM_FR,
                                   A_CLAY_MI,A_SAND_MI, A_SILT_MI, A_SOM_LOI,A_P_AL, A_P_WA, A_P_CC,
                                   A_PH_CC, A_CEC_CO,A_K_CO_PO, A_K_CC,
                                   M_GREEN = FALSE)]

# translate labels
dt[B_SOILTYPE_AGR == 'dekzand', soiltype := "sand"]
dt[B_SOILTYPE_AGR == 'veen', soiltype := "peat"]
dt[B_SOILTYPE_AGR == 'zeeklei', soiltype := "clay"]

dt[B_LU_BRP == 256, cropname := "sugarbeet"]
dt[B_LU_BRP == 259, cropname := "silage maize"]
dt[B_LU_BRP == 265, cropname := "permanent grassland"]
dt[B_LU_BRP == 2014, cropname := "ware potato"]
dt[,cropname := factor(cropname,levels = c("sugarbeet", "silage maize", "permanent grassland", "ware potato"))]

ggplot(
  data = dt,
  mapping = aes(y = i_gw_nlea,
                x = cropname,
                fill = cropname)
) +
  geom_col() +
  facet_wrap(facets = "soiltype", ncol = 1) +
  ylab("N leaching risk (i_gw_nlea)") +
  xlab("") +
  coord_flip() +
  scale_fill_viridis_d() +
  theme(legend.position = "none")
```

## Contaminant buffering
BLN has one function to score the risk of groundwater pollution with pesticides.
This function, `bln_wat_pesticide` estimates the likelihood that a persistent
pesticide penetrates in the soil below 1 metres. Because this function scores a
risk, a poor score does not necessarily mean that groundwater is or will be 
contaminated as the function does not take management or actual precipitation 
into account. The function uses `BLN::bln_calc_psp` to determine a precipitation
surplus, this depends on the cultivated crop and whether green manure was grown. 
The other driving variable for this indicator is soil organic matter content.
@vandenDool2021 describe the function in more detail and describe a demonstration 
in practice. They based the development of the function on the formula described
by @Tiktak2006.

```{r Pesticide demo, fig.height=4, fig.width=7}

dt.dz <- data.table(
  ID = 15,
  B_LU_BRP = 233,
  B_SOILTYPE_AGR = 'dekzand',
  soiltype = "sand",
  A_SOM_LOI = seq(1.2, 8.2, 0.2),
  A_CLAY_MI = 20,
  A_SAND_MI = 15,
  A_SILT_MI = 10,
  M_PESTICIDES_DST = FALSE,
  M_MECHWEEDS = FALSE,
  M_GREEN = FALSE
  )

dt.zk <- data.table(
  ID = 15,
  B_LU_BRP = 233,
  B_SOILTYPE_AGR = 'zeeklei',
  soiltype = "marine clay",
  A_SOM_LOI = seq(1.0, 8.0, 0.2),
  A_CLAY_MI = 20,
  A_SAND_MI = 15,
  A_SILT_MI = 10,
  M_PESTICIDES_DST = FALSE,
  M_MECHWEEDS = FALSE,
  M_GREEN = FALSE
  )

dt.ve <- data.table(
  ID = 15,
  B_LU_BRP = 233,
  B_SOILTYPE_AGR = 'veen',
  soiltype = "peat",
  A_SOM_LOI = seq(1.9, 19.9, 0.2),
  A_CLAY_MI = 20,
  A_SAND_MI = 15,
  A_SILT_MI = 10,
  M_PESTICIDES_DST = FALSE,
  M_MECHWEEDS = FALSE,
  M_GREEN = FALSE
  )

dt.da <- data.table(
  ID = 15,
  B_LU_BRP = 233,
  B_SOILTYPE_AGR = 'dekzand',
  soiltype = "peaty sand",
  A_SOM_LOI = seq(2.1, 19.7, 0.2),
  A_CLAY_MI = 20,
  A_SAND_MI = 15,
  A_SILT_MI = 10,
  M_PESTICIDES_DST = FALSE,
  M_MECHWEEDS = FALSE,
  M_GREEN = FALSE
  )

dt <- rbindlist(list(dt.dz, dt.zk, dt.ve, dt.da))

# groundwater quantity and quality: pesticide leaching (I_H_PEST,from OBI)
dt[, i_gw_pest := bln_wat_pesticide(ID,B_LU_BRP, B_SOILTYPE_AGR, A_CLAY_MI, A_SAND_MI, A_SILT_MI,
                                    A_SOM_LOI, M_GREEN, M_MECHWEEDS, M_PESTICIDES_DST)]

ggplot(
  data = dt[A_SOM_LOI <=15],
  mapping = aes(x = A_SOM_LOI, y = i_gw_pest, colour = soiltype)
) +
  geom_line(size = 1.5) +
  scale_colour_manual(values = palette.colors(palette = "Okabe-Ito")[2:6]) +
  xlab("Soil organic matter, %") +
  ylab("Pesticide risk score") +
  ylim(0,1) +
  xlim(0,15)
  

```

## References
